Minimal trap spaces of Logical models are maximal siphons of their Petri net encoding

### 2022-05-12 cmsb22.tex
Some words in the conclusion mostly

\begin{table}[!htb]
  \caption{Timing comparisons between the proposed method (i.e., Trappist) and PyBoolNet on the PyBoolNet's repository.
    Column \(n\) denotes the number of nodes of each model, Column \(|M|\) denotes the number of minimal trap spaces and for each method is given the computation time in seconds.
    A number in bold indicates a ratio greater than three.}
  \centering
  \label{tab:pyboolnet_repo}
  \begin{tabular}{lrrrrrr}
    \toprule
    model & $n$ & $|M|$ & PyBoolNet & Trappist \\ \midrule
    arellano\_rootstem & 9 & 4 & 0.04 & \textbf{0.31}\\
    calzone\_cellfate & 28 & 27 & 0.03 & 0.08\\
    dahlhaus\_neuroplastoma & 23 & 32 & 0.04 & 0.11\\
    davidich\_yeast & 10 & 12 & 0.03 & 0.08\\
    dinwoodie\_life & 15 & 7 & 0.03 & 0.08\\
    dinwoodie\_stomatal & 13 & 1 & 0.04 & 0.05\\
    faure\_cellcycle & 10 & 2 & 0.04 & 0.04\\
    grieco\_mapk & 53 & 18 & 0.04 & 0.08\\
    irons\_yeast & 18 & 1 & 0.05 & 0.04\\
    jaoude\_thdiff & 103 & >1000 & \textbf{1.73} & 0.40\\
    klamt\_tcr & 40 & 8 & 0.06 & 0.04\\
    krumsiek\_myeloid & 11 & 6 & 0.03 & 0.03\\
    multivalued & 13 & 4 & 0.02 & 0.02\\
    n12c5 & 11 & 5 & \textbf{34.37} & 0.06\\
    n3s1c1a & 2 & 2 & 0.02 & 0.03\\
    n3s1c1b & 2 & 2 & 0.02 & 0.03\\
    n5s3 & 4 & 3 & 0.03 & 0.03\\
    n6s1c2 & 5 & 3 & 0.03 & 0.04\\
    n7s3 & 6 & 3 & 0.02 & 0.02\\
    raf & 3 & 2 & 0.03 & 0.02\\
    randomnet\_n15k3 & 15 & 3 & 0.04 & 0.05\\
    randomnet\_n7k3 & 7 & 10 & 0.03 & \textbf{0.19}\\
    remy\_tumorigenesis & 34 & 25 & \textbf{1.79} & 0.13\\
    saadatpour\_guardcell & 13 & 1 & 0.02 & 0.03\\
    selvaggio\_emt & 56 & >1000 & \textbf{1.03} & 0.23\\
    tournier\_apoptosis & 12 & 3 & 0.05 & 0.06\\
    xiao\_wnt5a & 7 & 4 & 0.06 & 0.03\\
    zhang\_tlgl & 60 & 156 & 0.26 & 0.17\\
    zhang\_tlgl\_v2 & 60 & 258 & 0.19 & 0.17\\
    \bottomrule
  \end{tabular}
\end{table}

As shown in Table~\ref{tab:pyboolnet_repo}, for most of the models of the official PyBoolNet repository\footnote{\url{https://github.com/hklarner/pyboolnet/tree/master/pyboolnet/repository}}, the results are comparable with all minimal trap spaces found very fast.
Note however that on some very small models, our proposed method was sometimes slower than PyBoolNet, but still way under one second.
Moreover, we believe that the result on the arellano\_rootstem model is caused by the cold start of the JVM for BioLQM.
On the contrary, on every model that was a bit challenging for PyBoolNet, the new method was far more efficient with speedups between one and two orders of magnitude.

@@

We used a set of real-world Boolean models lying in various scales collected from various bibliographic sources.
These models are quite big (in size), complex (i.e., having high average in-degree, which is related to the number of prime-implicants) and most of them have never been fully analyzed.
We then applied the proposed method and PyBoolNet to computing minimal trap spaces of these real-world models.
It is notable that unlike existing analysis shown in the literature, we did not fix specific values for source nodes (i.e., some node \(v\) such that \(f_v = v\)) in these models.
Table~\ref{tab:result_real} shows the experimental results on real-world models. Hereafter, we analyze in detail the results with respect to minimal trap space computation.

@@

Though we benchmarked this new approach against the state-of-the-art tool PyBoolNet, there are many more evaluations that we plan to do in the future.
First, the BioLQM platform that we are using is providing another implicant-based method using BDDs in \url{http://colomoto.org/biolqm/doc/tools-trapspace.html} and though we expect it to behave mostly like PyBoolNet, that remains to be checked.
Note that this also raises the question of replacing altogether the BioLQM preprocessing step we use to obtain the Petri net encoding, since its use of BDDs might not be optimal for that step.
The trade-off between a small Petri net, and hence small ASP, and the time it takes to compute it has to be evaluated in depth.
Second, with the experimental results shown in this paper, it is not enough information to distinctly expose the practical differences between our ASP encoding and that of PyBoolNet.
Hence, we plan to conduct experiments on more real-world models or maybe random models that can be randomly generated by using BoolNet~\cite{mussel2010boolnet}.

@@

However, the current method already appears to perform very well even on the biggest models we have considered.
We thus plan to propose its inclusion in the CoLoMoTo effort to make it readily available to modellers.

>>>>

\begin{table}[!htb]
  \caption{Timing comparisons between the proposed method (i.e., Trappist) and PyBoolNet on the PyBoolNet repository.
    Column \(n\) denotes the number of nodes of each model, Column \(|M|\) denotes the number of minimal trap spaces and for each method is given the computation time in seconds, asking only for the first 1000 trap spaces.
    A number in bold indicates a ratio greater than three.}
  \centering
  \label{tab:pyboolnet_repo}
  \begin{tabular}{lrrrrrr}
    \toprule
    model & $n$ & $|M|$ & PyBoolNet & Trappist \\ \midrule
    arellano\_rootstem & 9 & 4 & 0.04 & \textbf{0.31}\\
    calzone\_cellfate & 28 & 27 & 0.03 & 0.08\\
    dahlhaus\_neuroplastoma & 23 & 32 & 0.04 & 0.11\\
    davidich\_yeast & 10 & 12 & 0.03 & 0.08\\
    dinwoodie\_life & 15 & 7 & 0.03 & 0.08\\
    dinwoodie\_stomatal & 13 & 1 & 0.04 & 0.05\\
    faure\_cellcycle & 10 & 2 & 0.04 & 0.04\\
    grieco\_mapk & 53 & 18 & 0.04 & 0.08\\
    irons\_yeast & 18 & 1 & 0.05 & 0.04\\
    jaoude\_thdiff & 103 & >1000 & \textbf{1.73} & 0.40\\
    klamt\_tcr & 40 & 8 & 0.06 & 0.04\\
    krumsiek\_myeloid & 11 & 6 & 0.03 & 0.03\\
    multivalued & 13 & 4 & 0.02 & 0.02\\
    n12c5 & 11 & 5 & \textbf{34.37} & 0.06\\
    n3s1c1a & 2 & 2 & 0.02 & 0.03\\
    n3s1c1b & 2 & 2 & 0.02 & 0.03\\
    n5s3 & 4 & 3 & 0.03 & 0.03\\
    n6s1c2 & 5 & 3 & 0.03 & 0.04\\
    n7s3 & 6 & 3 & 0.02 & 0.02\\
    raf & 3 & 2 & 0.03 & 0.02\\
    randomnet\_n15k3 & 15 & 3 & 0.04 & 0.05\\
    randomnet\_n7k3 & 7 & 10 & 0.03 & \textbf{0.19}\\
    remy\_tumorigenesis & 34 & 25 & \textbf{1.79} & 0.13\\
    saadatpour\_guardcell & 13 & 1 & 0.02 & 0.03\\
    selvaggio\_emt & 56 & >1000 & \textbf{1.03} & 0.23\\
    tournier\_apoptosis & 12 & 3 & 0.05 & 0.06\\
    xiao\_wnt5a & 7 & 4 & 0.06 & 0.03\\
    zhang\_tlgl & 60 & 156 & 0.26 & 0.17\\
    zhang\_tlgl\_v2 & 60 & 258 & 0.19 & 0.17\\
    \bottomrule
  \end{tabular}
\end{table}

As shown in Table~\ref{tab:pyboolnet_repo}, for most of the models of the official PyBoolNet repository\footnote{\url{https://github.com/hklarner/pyboolnet/tree/master/pyboolnet/repository}}, the results are comparable with all minimal trap spaces found very fast.
Note however that on some very small models, our proposed method is sometimes slower than PyBoolNet, but still significantly under one second.
Moreover, we believe that the result on the arellano\_rootstem model is caused by the cold start of the JVM for BioLQM.
On the contrary, on every model that was a bit challenging for PyBoolNet, the new method is far more efficient with speedups between one and two orders of magnitude.

@@

We used a set of real-world Boolean models lying in various scales collected from various bibliographic sources.
These models are quite big (in size), complex (i.e., having high average in-degree, which is related to the number of prime-implicants) and most of them have never been fully analyzed.
We then applied the proposed method and PyBoolNet to computing minimal trap spaces of these real-world models.
It is notable that unlike existing analysis shown in the literature, we did not fix specific values for source nodes (i.e., some node \(v\) such that \(f_v = v\)) in these models.
Table~\ref{tab:result_real} shows the experimental results on those models. Hereafter, we analyze in detail the results with respect to minimal trap space computation.

@@

Though we benchmarked this new approach against the state-of-the-art tool PyBoolNet, there are many more evaluations that we plan to do in the future.
First, the BioLQM platform that we are using is providing another implicant-based method using BDDs in \url{http://colomoto.org/biolqm/doc/tools-trapspace.html} and though we expect it to behave mostly like PyBoolNet, that remains to be checked.
Note that this also raises the question of replacing altogether the BioLQM preprocessing step we use to obtain the Petri net encoding, since its use of BDDs might not be optimal for that step.
The trade-off between a small Petri net, and hence small ASP, and the time it takes to compute it (in other words, the tradeoff of allowing redundant constraints) has to be evaluated in depth.

Second, the experimental results shown in this paper mostly expose the differences caused by the prime-implicant phase of PyBoolNet. There is much more information required to distinctly study the practical differences between our ASP encoding and that of PyBoolNet, and notably their size/efficiency ratio.
Hence, we plan to conduct experiments on more real-world models or maybe random models that can be randomly generated by using BoolNet~\cite{mussel2010boolnet}.
Such experiments should include a time-course of the number of ASP solutions found for proper comparison of the APS encodings.

@@

However, the current method already appears to already perform very well even on the biggest models we have considered.
We thus plan to propose its inclusion in the CoLoMoTo effort to make it readily available to modellers.
